Sensor Clusterization in D-optimal Design in Infinite Dimensional Bayesian Linear Inverse Problems

### 2020-11-22 paper.tex
Ready for new bound paper

\title[Sensor Clusterization in D-optimal design in infinite
  dimensions]{Sensor Clusterization in D-optimal Design in Infinite
  Dimensional Bayesian Inverse Problems}

@@

\begin{abstract}
  We investigate the problem of sensor clusterization in optimal
  experimental design for infinite-dimensional Bayesian inverse
  problems. We suggest an analytically tractable model for such
  designs and reason how it may lead to sensor clusterization in the
  case of iid measurement noise. We also show that in the case of
  spatially correlated measurement error clusterization does not
  occur. As a part of the analysis we prove a matrix determinant
  lemma analog in infinite dimensions, as well as a lemma for
  calculating derivatives of $\log \det$ of operators.
\end{abstract}

@@

Designing experiments is usually done by optimizing some \emph{design
  criterion} and this is true both for frequentists
\cite{Silvey13,Ucinski05} as well as for Bayesians
\cite{ChalonerVerdinelli95}, see \cite{ChalonerVerdinelli95} for an
investigation of the analogy between the two approaches. Although
there is a plethora of design criteria, in the current work we will
focus on just one of these, commonly referred to as \emph{D-optimal
  design}. It has a simple and appealing motivation in the Bayesian
context as explained in \cite{ChalonerVerdinelli95}: for a linear
model, a D-optimal design seeks to maximize the expected information
gain (KL divergence \cite{KullbackLeibler51,CoverThomas12}) between
posterior and prior. In finite dimensions, maximizing the expected
information gain amounts to minimizing the determinant of the
posterior covariance matrix. In a frequentist settings, an optimal
design minimizes the volume of the uncertainty ellipsoid \cite[page
  16]{Ucinski05}, but this is done for the Fisher information matrix
and not the posterior covariance. However, \cite{ChalonerVerdinelli95}
shows that the latter is just a regularized version of the former.

The previous discussion is classical for experimental design when
inference is to take place over a finite (not too large) number of
parameters. The subject of optimal experimental design for function
inference in a Bayesian context was pioneered by
\cite{AlexanderianGloorGhattas14,AlexanderianPetraStadlerEtAl16,
  AlexanderianPetraStadlerEtAl14}. Similarly to the finite dimensional
case, it can be shown that a D-optimal design arises naturally for
linear models when one wishes to maximize the KL divergence between
posterior and prior and this amounts to minimizing the determinant of
the posterior covariance operator (understood as a product of its
eigenvalues). Some difficulties arise in the process, but remedies can
be found as shown in \cite{AlexanderianGloorGhattas14}.

It seems counter intuitive that when one computes an optimal design
using the D-optimal criterion (and others that will not be
investigated here), the optimization process results in measurements
that are very similar. For example, if a measurement is thought of as
measuring some function value at $\x \in \Omega \subseteq \R^d,
d=1,2,3$ (with added error) then the optimization procedure sometimes
places sensors in close proximity to each other (as can be seen
in figure \ref{fig:clusterization illustration}). Following
\cite{Ucinski05}, we refer to this phenomenon as \emph{sensor
  clusterization}.

\subsection{Related Work}
The phenomenon of sensor clusterization seems to be known in several
different contexts. In a frequentist and finite-dimensional context,
\cite{Fedorov96} and \cite[chapter 2.4.3]{Ucinski05} discuss this
phenomenon and suggest an approach called clusterization-free design,
where the user enforces measurement locations to be far from each
other. One way to do this is by introducing correlated errors which,
philosophically, accounts for both measurement error and model error.
Another method considered is imposing distance constraints between
measurements. A somewhat different approach is suggested in \cite[page
  49]{Fedorov12}, where close by design measurements are merged --- a
procedure which obviously does not avoid clusterization. The same problem
arises in time-series analysis for pharmacokinetic experiments. The
authors of \cite{Hooker09} suggest modeling auto-correlation time in
the noise model, which is equivalent to the correlated errors
mentioned above.

@@

\subsection{Contribution}
We propose and thoroughly study an analytically tractable model for
understanding D-optimal designs. In this model, an optimal design
arises as a solution to a constrained optimization problem, formulated
using Lagrange multipliers (section \ref{section:D and grad}). In the
case of no model error we show this problem is in fact a set of
eigenvector-eigenvalue problems (equation \eqref{eq:eigenproblem}). We
then characterize an optimal design as one that reduces the highest
uncertainties, and show how that can give rise to repeated
measurements (section \ref{subsec:clusterization} and figure
\ref{fig:clusterization}).

In the process we generalize several lemmas from linear algebra to
infinite-dimensional settings --- the Matrix Determinant Lemma and a
lemma for calculating the derivative of determinants. We also provide
other tools for understanding D-optimal designs in
infinite-dimensional Bayesian inverse problems, including lemmas for
calculating the increase in the design criterion per observation ---
Lemmas \ref{lemma:design increase} and Corollary \ref{cor:zero mod
  err}. %%  We also show that in our model with independent measurement
%% noise, an interesting phenomenon arises, in which the eigenvectors
%% of the measurement operator (denoted below $\obs^*\obs$) are the
%% same as the eigenvectors of the prior covariance in observation
%% space, which helps arrive at the results outlined in the previous
%% paragraph.
The arguments presented are completely generic and apply equally to
any of the scenarios mentioned in the introduction.

%% \subsection{Limitations}\label{subsec:limitations}
%% There are two main drawbacks of the arguments presented here. First,
%% one can argue that we allow considerably less restrictive observations
%% than any real-life problem does. This is because our observation model
%% allows taking measurements that are arbitrarily weighted linear
%% combinations of eigenvectors. The second drawback is that we do not
%% show rigorously that clusterization necessarily occurs in our model,
%% we only show that it can occur.

\subsection{An Example of Clusterization}\label{subsec:example}
\begin{figure}
  \begin{tikzpicture}[thick, scale=1.3, every node/.style={scale=0.99}]
    \begin{axis}
      [
      xmin = 0,
      xmax = 3.14,
      xlabel = {$x$},
      ylabel = posterior std,
      ymin   = 0,
      %compat = 1.3,
      % ymax   = 130,
      % ytick = \empty,
      legend cell align=left,
      % legend style={at={(0.45,0.2)}}
      legend pos= outer north east 
      ]
      % \draw[black!30!white, thin] (50,0) -- (50,130);
      % 
      %% \addplot [thin, black, mark=none] table{stdv-heat-sens1-var1.txt};
      %% \addlegendentry{1 sensors};

@@

      %% \addplot [black,  only marks, mark=x, mark size=1.5] 
      %% table{locs-heat-sens1-var1.txt}; 
      %% \addplot [blue,   only marks, mark=x, mark size=1.5]
      %% table{locs-heat-sens2-var1.txt}; 
      %% \addplot [red,    only marks, mark=x, mark size=1.5]
      %% table{locs-heat-sens3-var1.txt};
      \addplot [green,  only marks, mark=x, mark size=1.5] 
      table{locs-heat-sens4-var1.txt}; 
      \addplot [purple, only marks, mark=x, mark size=1.5] 
      table{locs-heat-sens5-var1.txt}; 
      \addplot [cyan,   only marks, mark=x, mark size=1.5] 
      table{locs-heat-sens6-var1.txt};

    \end{axis}
  \end{tikzpicture}
  \caption{Shown are pointwise posterior standard deviations for
    various numbers of sensors, and location of these sensors for the
    1D heat equation with homogeneous Dirichlet boundary described in
    section \ref{subsec:example}. Note the clusterization that occurs
    when 6 sensors are used --- only 4 measurement locations are
    included. Observed data does not influence the pointwise
    variances, hence it is omitted.}
\end{figure}\label{fig:clusterization illustration}
In section \ref{section:abstract OED} we present a more abstract and
general formulation of the problem, but for the purpose of
illustration, we present the problem via a toy model --- the 1D heat
equation in $[0,\pi]$ with homogeneous Dirichlet boundary
conditions.

@@

We would like to infer the initial condition $u_0$. For that purpose,
we measure $u$ at some set of locations $\x_j \in [0,\pi], j=1, \dots,m$
and a final time $T > 0$. We assume centered Gaussian measurement
error, so we can observe $v(\x_j,T) = u(\x_j,T) + \eps(\x_j)$ with
$\eps(\x_j) \sim \normal(0, \sigma^2), \sigma > 0$ iid.  We model the
initial condition as $u_0 \sim \normal(0,\prcov)$, for $\prcov =
(-\Delta)^{-1}$ with homogeneous Dirichlet boundary condition. It is
well known \cite{Tarantola05} that for linear problems, with Gaussian
prior and error, the posterior is also Gaussian with a covariance that
does not depend on the observed data. The posterior covariance
$\postcov$ is known to have a closed form formula, even in infinite
dimensions\cite{Stuart10}. If we denote by $\fwd$ the dynamics
operator, so that $u( \cdot,T) = \fwd u_0$, and the observation
operator $\obs$ so that $u(\x_j,T) = (\obs u)_j, j=1,\dots,m$, then the
posterior covariance is known and depends on $\prcov, \fwd, \obs$ and
$\sigma^2$ (see section \ref{section:abstract OED} and
\eqref{eq:postcov} specifically).

We will consider generalization of the information-theoretic design
criterion presented in section \ref{section:OED intro} to infinite
dimensions (section \ref{subsec:D optimal design}). We choose
$\x_j,j=1,\dots,m$ so as to minimize an expression analogous to the
log-determinant of the posterior covariance operator (corresponding to
a D-optimal design). We can see the clusterization effect in figure
\ref{fig:clusterization illustration} --- the design criterion
suggests placing sensors in locations that are extremely close.

\subsection{Remarks to the Reader}\label{subsec:plan}
The Lagrange multipliers problem for iid measurement noise is
discussed in section \ref{section:vanishing}. In that section, it is
shown how sensor clusterization can occur. The reader may skip if they
are constrained by time. It is advised that a reader less comfortable
with Hilbert and Banach spaces mostly ignore these terms and consider
$\hil$ and $\ban$ finite dimensional vector spaces.

%% Seeking an optimal design amounts to maximizing $\tar(\obs)$ in
%% $\obs$, subject to constraints posed on $\obs$. These enter as
%% constraints on the allowed set of measurement vectors we can choose
%% from --- the rows of $\obs$. These constraints are determined by the
%% problem structure --- for example, in the 1D heat equation example
%% considered in section \ref{subsec:example}, we were constrained to
%% choose $\obs = (\meas_1,\dots,\meas_m)^t$ with $\meas_j = \delta_{\x_j}$
%% for some $\x_j \in [0,\pi]$. We present an approach to understanding
%% optimal experiment design based on the first-order necessary condition
%% for optimal design, where the constraints are incorporated using
%% Lagrange multipliers. This condition is used to understand what causes
%% clusterization. One problem with such an approach is that the set we
%% can choose $\obs$ from is complicated and difficult to work with. For
%% example, if $\ban = C(\Omega)$ for some compact $\Omega$, then the
%% allowed set of measurement vectors may only be point evaluations --- a
%% set that is not easy to describe as a level set of some simple
%% function, a description crucial for the use of Lagrange multipliers.
%% In order to facilitate analysis, we consider an easier constraint in
%% section \ref{subsec:unit length}. Reverting to this simpler constraint
%% is a limitation of our analysis but it provides some insight into the
%% cause of clusterization.

@@

The theoretical foundations for inverse problems over function spaces
can be found in \cite{Stuart10}. This section is mostly devoted to
setting notation. At this point, it is useful to record a known result
regarding the posterior covariance operator of our inverse problem,
namely
\begin{align}\label{eq:postcov precursor}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd )^{-1}.
\end{align}

\subsection{Bayesian Inverse Problems}\label{section:abstract OED}
Let $\hil$ be a Hilbert space and $\ban$ a Banach space, both
infinite-dimensional. Consider $\fwd: \hil \to \ban$, a linear
operator (the ``forward operator''). We are interested in forward
operators that are strongly smoothing (have fast decaying singular
values --- the heat operator from section \ref{subsec:example} is a
prime example).

We want to infer $\param \in \hil$, some parameter of the dynamics
given noisy observations of $\fwd \param$. We model $\param \sim
\normal(0,\prcov)$ with some appropriate covariance operator $\prcov$
\cite{Stuart10}. Observations (measurements) are taken via the linear
observation operator $\obs \in ( \ban^* )^m$, where $m$ is the number
of observations allowed (e.g. sensors at our disposal). In an analogy
with linear algebra (where row vectors are thought of as linear
functionals), we think of $\obs$ as having ``rows'' $\meas_j,
j=1,\dots,m$:
\begin{equation}\label{eq:O}
  \obs = (\meas_1,\dots, \meas_m)^t, \meas_j \in \ban^*, j = 1,\dots,m.
\end{equation}
This way, for $u \in \ban$ we have $\obs u = (\meas_1(u), \dots,
\meas_m(u) )^t \in \R^m$.
%% but we may drop the parentheses and write $\meas_j u = \meas_j(u)$.
A few observations regarding $\obs$ are in order. First, it is good to
keep in mind that $( \ban^* )^m$ is a Banach space with norm $\| \obs
\| = \sum_{j=1}^m \|\meas_j\|$. Next, for $u \in \ban$ and $v\in
\R^m$:
\begin{align*}
  \big (\obs^*v \big ) (u) &= \langle v, \obs u \rangle_{\R^m} = \sum_{j=1}^m  v_j \meas_j(u)
  = v^t \left ( \obs u \right ) = (v^t \obs) (u),
\end{align*}
and thus:
\begin{align}\label{eq:obs*}
  \obs^*v &= \sum_{j=1}^m v_j \meas_j = v^t \obs.
\end{align}

\begin{observation}
  It is best to think of $\meas_j,j=1,\dots,m$ as row vectors and of
  $\obs$ as a matrix with rows $\meas_j$. 
\end{observation}

Each measurement consists of a linear functional, chosen from an
allowed set of linear functionals in $\ban^{*}$. Data is acquired via
noisy measurements
\begin{align*}
  \data := \obs (\fwd \param + \eps') + \eps = \obs \fwd \param + \obs \eps' + \eps
\end{align*}
with $\data \in \R^m$ and $\eps, \eps'$ defined next. Measurement
error consists of two parts. First, there is spatially correlated
error $\eps' \sim \normal(0,\modcov)$, modeled as a centered Gaussian
measure on $\ban$ with covariance operator $\modcov: \ban^* \to \ban$
(see \cite[section 6]{Stuart10} for some details on Gaussian measures
on Banach spaces). Then there is measurement / independent error
$\eps \sim \normal(0, \sigma^2 I_m)$, with $I_m$ the $m \times m$
identity. Both error terms and the prior are assumed independent of
each other. Let us now understand the terms present in the above
expression.

\subsubsection{Dynamics and Observation Operators}\label{subsec:dynamics}
One can think of $(\obs \fwd \param)_j$ as pointwise evaluations of a
continuous function if $\ban$ is some appropriate function space,
e.g.\ $C(\Omega)$, the Banach space of real-valued continuous
functions over a compact $\Omega$ with the $\| \cdot \|_{\infty}$
norm. These measurements can be represented using the linear
functionals $\delta_{\x_j} \in C(\Omega)^*$. In a different setting,
we may be able to measure other quantities, e.g.\ some Fourier
coefficients as is the case in MRI applications. Either way, the rows
of $\obs$ are linear functionals and we refer to these as {\it
  measurement vector}s. Note that we cannot choose every measurement
vector we want, though. For example, we may be restricted only to
pointwise evaluations of $\fwd \param$ by the sensors at our disposal,
in which case measuring the mean $\ell(u) = \int_{\Omega}u$ (or any
non-atomic measure) of a function is not possible.

\subsubsection{Error Terms}
Considering the sum of the error terms it is easy to see that $\obs
\eps' + \eps \in \R^m$. It is a centered Gaussian random vector and
its covariance matrix is
\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps)  (\obs \eps' + \eps)^t ] 
    % 
    % 
    = \obs \modcov \obs^* + \sigma^2I , 
  \end{split}
\end{align}
where
\begin{equation}\label{eq:modcov explained}
  [\obs \modcov \obs^*]_{ij} = e_i^t \obs \modcov \obs^* e_j = \meas_i (\modcov \meas_j),
\end{equation}
where the first equality in \eqref{eq:modcov explained} holds by
definition and the second by \eqref{eq:obs*}. The explicit dependence
on $\obs$ will be mostly dropped for notational convenience, so
$\Sigma(\obs) = \Sigma$. Thus, for a fixed $\obs$ (i.e.\ a fixed set
of measurements) we can equivalently write $\data = \obs \fwd \param +
\bar{\eps}$ with $\bar{\eps} \sim \normal(0,\Sigma)$. At this point,
it is useful to rewrite the posterior covariance operator of our
inverse problem from \eqref{eq:postcov precursor}
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}

>>>>

\title[Sensor Clusterization in D-optimal design in infinite
  dimensions]{Sensor Clusterization in D-optimal Design in Infinite
  Dimensional Bayesian Linear Inverse Problems}

@@

\begin{abstract}
  We investigate the problem of sensor clusterization in D-optimal
  experimental design for infinite-dimensional Bayesian linear inverse
  problems. We suggest an analytically tractable model for D-optimal
  designs and reason how it may lead to sensor clusterization in the
  case of iid measurement noise. We also spatially correlated
  measurement and model errors mitigate the clusterization
  phenomenon. As a part of the analysis we prove a Matrix Determinant
  Lemma in Hilbert spaces and a lemma for calculating
  $\frac{\der}{\der t} \log \det (I + X(t))$ for certain
  operator-valued functions. We also show how to decompose $M = AA^t$
  with $A$ constrained to have unit norm columns.
\end{abstract}

@@

Designing experiments is usually done by optimizing some \emph{design
  criterion}. This is true both for frequentists
\cite{Silvey13,Ucinski05} as well as for Bayesians
\cite{ChalonerVerdinelli95}. See \cite{ChalonerVerdinelli95} for an
investigation of the analogy between the two approaches. Although
there is a plethora of design criteria, we focus on just one of these,
commonly referred to as \emph{D-optimal design}. It has a simple and
appealing motivation in the Bayesian context as explained in
\cite{ChalonerVerdinelli95}: Consider a linear model in finite
dimensions, with Gaussian prior and noise. In this model, maximizing
the expected information gain (KL divergence
\cite{KullbackLeibler51,CoverThomas12}) between posterior and prior
amounts to minimizing the determinant of the posterior covariance
matrix. In a frequentist setting, a D-optimal design minimizes the
volume of the uncertainty ellipsoid \cite[page 16]{Ucinski05}, but
this is done for the Fisher information matrix and not the posterior
covariance. However, \cite{ChalonerVerdinelli95} shows that the latter
is just a regularized version of the former.

The previous discussion is classical for experimental design when
inference takes place over a finite (not too large) number of
parameters. The subject of optimal experimental design for function
inference in a Bayesian context was pioneered by
\cite{AlexanderianGloorGhattas14, AlexanderianPetraStadlerEtAl16,
  AlexanderianPetraStadlerEtAl14}. Similarly to the finite dimensional
case, it can be shown that a D-optimal design arises naturally for
linear models when one wishes to maximize the KL divergence between
posterior and prior. This amounts to minimizing the determinant of the
posterior covariance operator (understood as a product of its
eigenvalues). Some difficulties arise in the process, but remedies can
be found as shown in \cite{AlexanderianGloorGhattas14}.

It seems counter intuitive that when one computes an optimal design
using the D criterion, the optimization process results in
measurements that are very similar. For example, if a measurement is
thought of as measuring some function value at $\x \in \Omega
\subseteq \R^d, d=1,2,3$ (with added error) then the optimization
procedure sometimes places sensors in very close proximity to each
other (as can be seen in figure \ref{fig:clusterization
  illustration}). Following \cite{Ucinski05}, we refer to this
phenomenon as \emph{sensor clusterization}.

\subsection{Related Work}
The phenomenon of sensor clusterization seems to be known in several
different contexts. In a frequentist and finite-dimensional context,
\cite{Fedorov96} and \cite[chapter 2.4.3]{Ucinski05} discuss this
phenomenon and suggest an approach called clusterization-free design.
In such designs, the user enforces measurement locations to be far
from each other. One way to do this is by introducing correlated
errors which, philosophically, accounts for both measurement error and
model error. Another method considered is imposing distance
constraints between measurements. A somewhat different approach is
suggested in \cite[page 49]{Fedorov12}, where close by design
measurements are merged --- a procedure which obviously does not avoid
clusterization. The same problem arises in time-series analysis for
pharmacokinetic experiments. The authors of \cite{Hooker09} suggest
modeling auto-correlation time in the noise model, which is equivalent
to the correlated errors mentioned above.

@@

Similarly, in the inverse problems community, work is mostly
computational and less theoretic. Model errors were considered in
\cite{Attia20, Koval20}. The former study is focused on inferring a
Quantity of Interest (QoI). The focus of the latter is reducing
forward solves, using randomized linear algebra. Both studies present
numerical techniques for finding optimal designs when model error is
present. Both are restricted to linear inverse problems (although in
the latter the authors use their method on a nonlinear problem by
taking a Laplace approximation for the posterior). Both find an
optimal design by first solving a continuous problem for sensor
weights. Said solution is then sparsified to give a binary
design. Both studies are successful in the task of Bayesian
inversion. However, neither of these studies mention any effect model
errors can have on sensor clusterization. The current study is mostly
theoretical and aims to fill the gap of understanding sensor
clusterization.

\subsection{Contribution}
We propose and thoroughly study a relaxed and analytically tractable
model for understanding D-optimal designs. In this model, D-optimal
design are solutions of a constrained optimization problem, formulated
using Lagrange multipliers (section \ref{section:D and grad}). Using
this model we rigorously show how model error mitigates clusterization
(section \ref{section:non vanishing}). We show how clusterization
occurs in a sequential design scenario (section
\ref{subsec:clusterization sequential}). Then, we consider a
simultaneous D-optimal design scenario (section
\ref{subsec:clusterization simultaneous}). We show that designs that
exhibit clusterization are just as optimal as clusterization free
designs.

A beautiful mathematical structure arises in D-optimal designs when no
model error is present. The Lagrange multipliers problems is in fact a
nonlinear eigenvalue problem for the observations and prior
covariance. The operator for which eigenvectors and eigenvalues are
sought is a sum of two operators. The first is the prior
covariance. The second is an outer product of the observations (see
section \ref{section:vanishing} for details and exact statement). This
structure helps us shed light on D-optimal designs. We characterize
these designs in Theorem \ref{thm:char}. One insight we prove is that
a D-optimal design reduces uncertainties where they are highest
first. Other interestig phenomena arise, but they require setting
notation, and are discussed later.

In the process we generalize several lemmas from linear algebra to
infinite-dimensional settings. We prove a Matrix Determinant Lemma in
\ref{lemma:MDL}. We generalize a lemma due to Lax \cite{lax97} for
calculating $\frac{\der}{\der t} \log \det (I + X(t))$, for an
operator valued function $X(t)$ in \ref{lemma:lax}. We also prove
lemmas in linear algebra. One constructs a decomposition $M = AA^t$
where $A$ has unit norm columns \ref{lemma:free}. Another shows
simultaneous diagonizability of the prior and outer product in the
nonlinear eigenvalue problem referenced above \ref{lemma:sim diag}. We
also provide other tools for understanding D-optimal designs in
infinite-dimensional Bayesian inverse problems. These includet tools
for calculating the increase in the design criterion per observation
--- Lemma \ref{lemma:design increase} and Corollary \ref{cor:zero mod
  err}.

\subsection{Limitations}\label{subsec:limitations}
There are two main drawbacks of the study presented here. Our relaxed
model does not consider any specific set of allowed
measurements. Rather, we take measurements in the unit ball in some
Hilbert space. This allows considerably less restrictive observations
than any real-life problem does. The second drawback is that we do not
show rigorously that clusterization necessarily occurs in a
simultaneous design. We only show that it is as reasonable as no
clusterization.

\subsection{An Example of Clusterization}\label{subsec:example}
\begin{figure}
  \begin{tikzpicture}[thick, scale=1.3, every node/.style={scale=0.99}]
    \begin{axis}
      [
      title={Posterior Pointwise Standard Deviations and D-Optimal Sensor Locations},  
      xmin = 0,
      xmax = 3.14,
      xlabel = {$x$},
      ylabel = posterior std,
      ymin   = 0,
      %compat = 1.3,
      % ymax   = 130,
      % ytick = \empty,
      legend cell align=left,
      % legend style={at={(0.45,0.2)}}
      legend pos= outer north east 
      ]
      % \draw[black!30!white, thin] (50,0) -- (50,130);
      % 
      %% \addplot [thin, black, mark=none] table{stdv-heat-sens1-var1.txt};
      %% \addlegendentry{1 sensors};

@@

      %% \addplot [black,  only marks, mark=x, mark size=1.5] 
      %% table{locs-heat-sens1-var1.txt}; 
      %% \addplot [blue,   only marks, mark=x, mark size=1.5]
      %% table{locs-heat-sens2-var1.txt}; 
      %% \addplot [red,    only marks, mark=x, mark size=1.5]
      %% table{locs-heat-sens3-var1.txt};
      \addplot [green,  only marks, mark=*, mark size=1.5] 
      table{locs-heat-sens4-var1.txt}; 
      \addplot [purple, only marks, mark=*, mark size=1.5] 
      table{locs-heat-sens5-var1.txt}; 
      \addplot [cyan,   only marks, mark=*, mark size=1.5] 
      table{locs-heat-sens6-var1.txt};

    \end{axis}
  \end{tikzpicture}
  \caption{The clusterization effect for the 1D heat equation
    described in section \ref{subsec:example}. Posterior pointwise
    standard deviations (lines) are plotted over the domain $[0,
      \pi]$, for varying numbers of sensors. Sensor locations
    (circles) were chosen to minimize (an expression analogous to) the
    determinant of the posterior covariance. The clusterization effect
    can be clearly seen for six sensors. Only four measurement
    locations are used --- two pairs of sensors are so close they are
    indistinguishable.}
  \label{fig:clusterization illustration}
\end{figure}

In section \ref{section:prelim} we present a more abstract and general
formulation of the inverse problem we consider. But, for the purpose
of illustration, we present clusterization via a toy model --- the 1D
heat equation in $[0,\pi]$ with a homogeneous Dirichlet boundary
condition.

@@

We would like to infer the initial condition $u_0$. For that purpose,
we measure $u$ at some set of locations $\x_j \in [0,\pi], j=1,
\dots,m$ and a final time $T > 0$. We assume centered Gaussian
measurement error, so we observe $v(\x_j,T) = u(\x_j,T) + \eps(\x_j)$
with $\eps(\x_j) \sim \normal(0, \sigma^2), \sigma > 0$ iid. We model
the initial condition as $u_0 \sim \normal(0,\prcov)$, for $\prcov =
(-\Delta)^{-1}$ with a homogeneous Dirichlet boundary condition. It is
well known \cite{Tarantola05} that for linear problems, with Gaussian
prior and error, the posterior is also Gaussian with a covariance that
does not depend on the observed data. The posterior covariance
$\postcov$ is known to have a closed form formula, even in infinite
dimensions\cite{Stuart10}. We denote by $\fwd$ the dynamics operator,
so that $u( \cdot,T) = \fwd u_0$, and the observation operator $\obs$
so that $u(\x_j,T) = (\obs u)_j, j=1,\dots,m$. The posterior
covariance is known and depends only on $\prcov, \fwd, \obs$ and
$\sigma^2$ (see section \ref{section:prelim} and \eqref{eq:postcov}
specifically).

We consider generalization of the information-theoretic design
criterion presented in the introduction to infinite dimensions
(section \ref{subsec:D optimal design} below). We choose
$\x_j,j=1,\dots,m$ to minimize (an expression analogous to) the
determinant of the posterior covariance operator. We will see later
how this corresponds to maximizing expected information gain.

The clusterization effect is illustrated in figure
\ref{fig:clusterization illustration}. Posterior pointwise standard
deviations are plotted over the domain $[0, \pi]$. Since the posterior
covariance does not depend on data, the plot has no reference to
actual data observed. The posterior covariance does, however, depend
on location of the measurement taken. In figure
\ref{fig:clusterization illustration}, measurement locations are
marked by circles. These were chosen to minimize (an expression
analogous to) the determinant of the posterior covariance. The
clusterization effect can be clearly seen for six sensors. It looks
like only four measurements were taken. The reason is that two pairs
of sensors are so close they are indistinguishable.

@@

\subsection{Overview}
In this section we define notations that will be used throughout the
paper. The theoretical foundations for inverse problems over function spaces
can be found in \cite{Stuart10}.

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let $\hilp$ and $\hilo$ be separable Hilbert spaces (the subscripts p
and o are for ``parameter'' and ``observation'', respectfully). We
denote $\hilo^*$ the Hilbert space of all linear functionals on
$\hilo$. Consider $\fwd: \hilp \to \hilo$, a linear operator (the
``forward operator''). We are interested in forward operators that are
strongly smoothing (have fast decaying modes --- the heat operator
from section \ref{subsec:example} is a prime example). Our goal is to
infer $\param \in \hilp$ --- some parameter of the dynamics --- given
noisy observations of $\fwd \param$. We take a Gaussian prior $\pr$
for $\param$. Hence, $\param \sim \pr = \normal(\prmean ,\prcov)$ with
some appropriate covariance operator $\prcov$ \cite{Stuart10}. It is
important to note that $\fwd \prcov \fwd^*$ is the prior covariance in
$\hilo$ \cite{Stuart10}. As such, it is assumed invertible --- an
assumption we rely on below. Throughout this paper, we denote $m$ the
number of measurements we are allowed to take. Observations
(measurements) are taken via the linear observation operator $\obs \in
( \hilo^* )^m$ (sometimes called a design). In an analogy with linear
algebra (where row vectors are thought of as linear functionals), we
think of $\obs$ as having ``rows'' $\meas_j, j=1,\dots,m$:
\begin{equation}\label{eq:O}
  \obs = (\meas_1,\dots, \meas_m)^t, \meas_j \in \hilo^*, j = 1,\dots,m.
\end{equation}
This way, for $u \in \hilo$ we have $\obs u = (\meas_1(u), \dots,
\meas_m(u) )^t \in \R^m$.
%% but we may drop the parentheses and write $\meas_j u = \meas_j(u)$.
%% A few observations regarding $\obs$ are in order.
%% First, it is good to keep in mind that $( \hilo^* )^m$ is a Hilbert
%% space with norm $\| \obs \| = \sum_{j=1}^m \|\meas_j\|$.
For $u \in \hilo$ and $v\in \R^m$:
\begin{align*}
  \big (\obs^*v \big ) (u) &= \langle v, \obs u \rangle_{\R^m} = \sum_{j=1}^m  v_j \meas_j(u)
  = v^t \left ( \obs u \right ) = (v^t \obs) (u),
\end{align*}
and thus:
\begin{align}\label{eq:obs*}
  \obs^*v &= \sum_{j=1}^m v_j \meas_j = v^t \obs.
\end{align}

%% \begin{observation}
%%   It is best to think of $\meas_j,j=1,\dots,m$ as row vectors and of
%%   $\obs$ as a matrix with rows $\meas_j$. 
%% \end{observation}

Each measurement is a linear functional, chosen from some subset of
$\hilo^{*}$. Data is acquired via noisy measurements
\begin{align}\label{eq:inverse problem}
  \data := \obs (\fwd \param + \eps') + \eps = \obs \fwd \param + \obs \eps' + \eps,
\end{align}
with $\data \in \R^m$ and $\eps, \eps'$ defined next. We consider two
types of noise / error terms. First, there is spatially correlated
model error $\eps' \sim \normal(0,\modcov)$, modeled as a centered
Gaussian measure on $\hilo$ with covariance operator $\modcov$. Then
there is measurement / independent error $\eps \sim \normal(0,
\sigma^2 I_m)$, with $I_m$ the $m \times m$ identity. Both error terms
and the prior are assumed independent of each other. The observation
and forward operators are merged $\tmp := \obs \fwd$. Since we need to
separate the observation operator $\obs$ from the forward operator
$\fwd$ we write this explicitly in \eqref{eq:inverse problem}. We now
turn to understand the terms present in \eqref{eq:inverse problem}.

\subsubsection{Dynamics and Observation Operators}\label{subsec:dynamics}
One can \textit{think} of $(\obs \fwd \param)_j$ as pointwise
evaluations of a continuous function\footnote{Of course, this is
  impossible if, e.g., $\hilo = L^2(\Omega)$ for some $\Omega$.}. Such
measurements can be represented using the linear functionals
$\delta_{\x_j} \in C(\Omega)^*$. In a different setting, we may be
able to measure other quantities, e.g.\ some Fourier coefficients as
is the case in MRI applications. Either way, the rows of $\obs$ are
linear functionals and we refer to these as {\it measurement
  vector}s. Note that we cannot choose every measurement vector we
want. For example, we may be restricted only to pointwise evaluations
of $\fwd \param$ by the sensors at our disposal, in which case
measuring, e.g., the mean $\ell(u) = \int_{\Omega}u$ of a function is
not possible.

\subsubsection{Error Terms}
Considering the sum of the error terms it is easy to see that
$\bar{\eps} := \obs \eps' + \eps \in \R^m$. It is a centered Gaussian
random vector and its covariance matrix is
\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps)  (\obs \eps' + \eps)^t ] 
    % 
    % 
    = \obs \modcov \obs^* + \sigma^2I , 
  \end{split}
\end{align}
where
\begin{equation}\label{eq:modcov explained}
  [\obs \modcov \obs^*]_{ij} = e_i^t \obs \modcov \obs^* e_j = \meas_i (\modcov \meas_j),
\end{equation}
with $e_j \in \R^m$ the $j$th standard basis vector. The first
equality in \eqref{eq:modcov explained} holds by definition and the
second by \eqref{eq:obs*}. The explicit dependence on $\obs$ will be
mostly dropped for notational convenience, so $\Sigma(\obs) =
\Sigma$. Thus, for a fixed $\obs$ (i.e.\ a fixed set of measurements)
we can equivalently write $\data = \obs \fwd \param + \bar{\eps}$ with
$\bar{\eps} \sim \normal(0,\Sigma)$. Taking $\modcov = 0$ is common
practice \cite{Tarantola05,KaipioSomersalo05,Vogel02}. This reduces to
the case where we take iid observations. Then $\Sigma = \sigma^2I$ is
simply a scalar matrix that does not depend on $\obs$. Note that
taking an error model with a non-scalar covariance as we do here
allows us to consider model error (modeled by $\modcov$) as well as
measurement error (modeled by $\sigma^2$). For example, say we believe
our forward model does not capture some small scale phenomenon.  Then
we may express this belief by saying $\tru = \fwd + \err$, with $\fwd$
depending on $\param$ and $\err$ depending on $\sspar$, with $\param
\perp \sspar$. We do not know much about this effect but it is
reasonable to assume it changes continuously in our domain. We (may
choose to) model it as $\normal (0, \modcov)$ and take $\modcov$ to
reflect the spatial (or other) variability we imagine $\err \sspar$
has. Such small scale phenomenon can arise as a modeling issue, where
we might not model the system in its entirety. It can also arise from
a numerical source, where our discretization of the system is not fine
enough to capture all small scale phenomena. In section
\ref{section:non vanishing}, we will see that assuming some
correlation in the error mitigates the clusterization phenomenon, as
reported in the literature \cite{Ucinski05}.

Finally, it is useful to record a known result regarding the posterior
covariance operator of our inverse problem:
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}

\subsection{D-Optimal Designs in Infinite Dimensions}\label{subsec:D optimal design} 
A D-optimal design maximizes expected KL divergence between posterior
and prior. It is useful to recall the definition of KL divergence for
an arbitrary prior measure:
$$
KL(\post||\pr) = \int \log \frac{\der \post}{\der \pr}(\param) \der \post(\param).
$$
